Density-dependent selection in evolutionary genetics: a lottery model of Grime's triangle

### 2016-10-28 manuscript.tex
latest

Nevertheless, there are strong indications there are broader principles governing the operation of selection. In many groups of organisms, including corals \citep{darling_2012}, insects \citep{southwood_1977}, fishes \citep{winemiller_1992}, zooplankton \citep{allan_76} and plants \citep{grime_1988}, different species can be divided into a small number of distinct trait clusters corresponding to fundamentally distinct ``primary strategies'' \citep{winemiller_2015}. The most famous example is Grime's plant trait classification scheme \citep{grime_1974,grime_1977,grime_1988}. Grime considered two broad determinants of population density: stress (persistent hardship e.g. due to resource scarcity, unfavorable temperatures or toxins) and disturbance (intermittent destruction of vegetation e.g. due to trampling, herbivory, pathogens, extreme weather or fire).  The extremes of these two factors define three primary strategies denoted by C/S/R respectively: competitors ``C'' excel in low stress, low disturbance environments; stress tolerators ``S'' excel in high stress, low disturbance environments; and ruderals ``R''  excel in low stress, high disturbance environments. Survival is not possible in high-stress, high-disturbance environments. Grime showed that measures of C, S and R across a wide range of plant species are anti-correlated, so that strong C-strategists are weak S and R strategists, and so on. Thus, plant species can be classified on a triangular C/S/R ternary plot \citep{grime_1974}. Trait classification schemes for other organisms closely parallel Grime's scheme \citep{winemiller_2015}.

@@

The alternative view is that stressful environments simply have a lower carrying capacity \citep{taylor_1990}; in our model, this means a greater per-individual territorial requirement represented by a lower $K$ for a given amount of space. In particular, it is argued that when stress is induced by a scarcity of consumable resources, competition for those resources would likely be intense. Thus, $b_i$ need not be particularly close to $d_i$, but contests for consumable resources at the juvenile phase would kill off most propagules before adulthood. In other words, the stressed population is a high density population where competition is important \citep{taylor_1990} (that is, high density relative to $K$, not relative to ideal conditions). We include both of these alternatives in our discussion here.

@@

\begin{figure}
\centering
\begin{tabular}{*{5}{c}}
  & Ideal & Disturbance* & Stress (G) & Stress (K) \\ \hline
  Parameter- & $d \ll 1$ & $d \approx 1$ & $b \ll 1$ & $b \ll 1$ \\
  regime & $b/d\gg 1$ & $b/d\gg 1$ & $b\approx d$ & $b>d$ \\
  Density $N/K$  & High & Low & Low & High \\
  $\Delta_+ n_i\propto$ & $b_i c_i$ & $b_i$ & $b_i$ & $b_i c_i$ \\
  Evolution for & $\uparrow b$, $ \uparrow c$ & $\uparrow b$, $\downarrow d$ & $\downarrow d$ & $\uparrow c$
\end{tabular}
\caption{\label{fig:table} The realization of Grime's environmental archetypes in our model, as well as the low-$K$ variant of the stress archetype. Shown are the mapping to our parameters of each archetype, the approximate dependence of $\Delta_+ n_i$ on $b_i$ and $c_i$, as well as the corresponding expected evolutionary changes in $b_i$, $c_i$ and $d_i$. *Mortality affects both adults and juveniles in the disturbance archetype, with $\Delta_+ n_i$ replaced by $(1-d_i)\Delta_+ n_i$ in Eq. \eqref{eq:delttot}.}
\end{figure}

@@

How does Fig. 4 compare to Grime's C/S/R strategies? In disturbed environments, we predict evolution for higher $b$ and lower $d$, but not higher $c$. Higher $b$ means higher fecundity, but not necessarily mass propagule production: $b$ represents only those propagules which sucessfully develop into juveniles in unoccupied territories. This is broadly consistent with the ruderal primary strategy. Plant ruderals devote a large proportion of their productivity to seed production \cite{grime_1977}, whereas the analogous ``opportunistic'' strategists in fishes have large intrinsic growth rates \citep{winemiller_1992}. In corals, a distinguishing feature of the ruderal cluster is brood spawning (rather than broadcast spawning), which corresponds to higher parental investment and lower overall propagule production, but may be important for successful propagation in low density, disturbed environments due to Allee effects \citep{darling_2012}. Lower $d$ could be achieved by improved individual resistance to physical destruction, but it is hard to reduce mortality in the face of severe disturbances. Given this constraint, shortening the time to reproductive maturity (the iteration time in our model) is an effective way of reducing the chance of death per iteration $d$. A short life cycle is probably the most consistent characteristic of ruderals \cite{grime_1977,winemiller_1992,darling_2012}.

In stressful environments, we predict evolution either for lower $d$ or higher $c$ depending on how the stress interpreted.

In ideal environments, we predict evolution for higher $b$ and $c$, but not lower $d$ (which is already small). In plants and corals, the ability to secure territories $c$ is strongly associated with rapid individual growth, and rapid individual growth is a defining trait of the competitor trait cluster \cite{grime_1977,winemiller_1992,darling_2012}.

>>>>

Nevertheless, there are strong indications there are broader principles governing the operation of selection. In many groups of organisms, including corals \citep{darling_2012}, insects \citep{southwood_1977}, fishes \citep{winemiller_1992}, zooplankton \citep{allan_76} and plants \citep{grime_1988}, different species can be divided into a small number of distinct trait clusters corresponding to fundamentally distinct ``primary strategies'' \citep{winemiller_2015}. The most famous example is Grime's plant trait classification scheme \citep{grime_1974,grime_1977,grime_1988}. Grime considered two broad determinants of population density: stress (persistent hardship e.g. due to resource scarcity, unfavorable temperatures or toxins) and disturbance (intermittent destruction of vegetation e.g. due to trampling, herbivory, pathogens, extreme weather or fire).  The extremes of these two factors define three primary strategies denoted by C/S/R respectively: competitors ``C'' excel in low stress, low disturbance environments; stress tolerators ``S'' excel in high stress, low disturbance environments; and ruderals ``R''  excel in low stress, high disturbance environments. Survival is not possible in high-stress, high-disturbance environments. Grime showed that measures of C, S and R across a wide range of plant species are anti-correlated, so that strong C-strategists are weak S and R strategists, and so on. Thus, plant species can be classified on a triangular C/S/R ternary plot \citep{grime_1974}. Trait classification schemes for other organisms are broadly analogous to Grime's scheme \citep{winemiller_2015}.

@@

The alternative view is that stressful environments simply have a lower carrying capacity \citep{taylor_1990}; in our model, this means a greater per-individual territorial requirement represented by a lower $K$ for a given amount of space. In particular, it is argued that when stress is induced by a scarcity of consumable resources, competition for those resources would likely be intense. Thus, $b_i$ need not be particularly close to $d_i$, but contests for consumable resources at the juvenile phase would kill off most propagules before adulthood. In other words, the stressed population is a high density population where competition is important \citep{taylor_1990} (that is, high density relative to $K$, not relative to ideal conditions).

@@

\begin{figure}
\centering
\begin{tabular}{*{5}{c}}
  & Ideal & Disturbance* & Stress (G) & Stress (K) \\ \hline
  Parameter- & $d \ll 1$ & $d \approx 1$ & $b \ll 1$ & $b \ll 1$ \\
  regime & $b/d\gg 1$ & $b/d\gg 1$ & $b\approx d$ & $b>d$ \\
  Density $N/K$  & High & Low & Low & High \\
  $\Delta_+ n_i\propto$ & $b_i c_i$ & $b_i$ & $b_i$ & $b_i c_i$ \\
  Evolution for & $\uparrow b$, $ \uparrow c$ & $\uparrow b$, $\downarrow d$ & $\downarrow d$ & $\uparrow c, \downarrow d$
\end{tabular}
\caption{\label{fig:table} The realization of Grime's environmental archetypes in our model, as well as the low-$K$ variant of the stress archetype. Shown are the mapping to our parameters of each archetype, the approximate dependence of $\Delta_+ n_i$ on $b_i$ and $c_i$, as well as the corresponding expected evolutionary changes in $b_i$, $c_i$ and $d_i$. *Mortality affects both adults and juveniles in the disturbance archetype, with $\Delta_+ n_i$ replaced by $(1-d_i)\Delta_+ n_i$ in Eq. \eqref{eq:delttot}.}
\end{figure}

@@

How does Fig. 4 compare to Grime's C/S/R strategies? In our comparison we will stick to fishes, corals and plants, for which three-way primary strategy schemes are well developed \citep{grime_1977,winemiller_1992,darling_2012}. The connection of our model to fish strategies is necessarily more tentative, given that fishes are motile and not all territorial and the starting assumption of our model is site-occupancy .

In disturbed environments, we predict evolution for higher $b$ and lower $d$, but not higher $c$. Higher $b$ means higher fecundity, but not necessarily mass propagule production: $b$ represents only those propagules which sucessfully develop into juveniles in unoccupied territories. This is broadly consistent with the ruderal primary strategy. Plant ruderals devote a large proportion of their productivity to seed production \cite{grime_1977}, whereas the analogous ``opportunistic'' strategists in fishes have large intrinsic growth rates \citep{winemiller_1992}. In corals, a distinguishing feature of the ruderal cluster is brood spawning (rather than broadcast spawning). This corresponds to higher parental investment and lower overall propagule production, but potentially also higher $b$ at low densities, since broadcast spawners are vulnerable to a powerful Allee effect at the egg fertilization stage \citep{knowlton_2001}. Lower $d$ could be achieved by improved individual resistance to physical destruction, but it is hard to reduce mortality in the face of severe disturbances. Given this constraint, shortening the time to reproductive maturity (the iteration time in our model) is an effective way of reducing the chance of death per iteration $d$. An exceptionally short life cycle is probably the most defining characteristic of ruderals \citep{grime_1977,winemiller_1992,darling_2012}.

In stressful environments, we predict evolution for lower $d$, and also  for higher $c$ in the low-$K$ interpretation of the stress archetype. Lowering $d$ is obviously essential when $b\ll 1$, and stress tolerant plants and corals have long life spans, allowing for long intervals between successful recruitments (and episodic broadcast spawning in corals). For fishes, the ``equilibrium'' strategy is the analogue of Grime's stress tolerator. This strategy is associated with resource limitation, and is also characterized by long life span, as well as high parental investment in tiny broods. This may reflect a high-$c$ strategy in the face of intense competition for severely limited resources (the low-$K$ interpretation).

In ideal environments, we predict evolution for higher $b$ and $c$, but not lower $d$ (which is already small). In plants and corals, the ability to secure territories $c$ is strongly associated with rapid individual growth, and rapid individual growth is a defining feature of the competitor trait cluster \cite{grime_1977,darling_2012}. Evolution for higher $b$ (which is closely related to the intrinsic low-density growth rate $r=b-d$ ) is not usually thought to be the expected outcome in high-density competitive environments. We will discuss this further below in the context of MacArthur's $r$/$K$ scheme. For now we simply note that the analogue of Grime's ideal environment for fishes occurs when resources limitations are not severe and the environment varies predictably over large temporal and spatial scales \cite{winemiller_1992,winemiller_2015}. The associated ``periodic'' strategy is characterized by enormous brood sizes, suggesting a strategy of ensuring that more propagules actually end up contesting areas favorable for development (higher $b$).

Indeed, the classic lottery model was originally applied to territorial reef fishes, and consequently the $l_i\gg 1$ assumption was
